---
title: "fiscal-demographic"
author: "Nick Gauthier"
date: "January 24, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=F, warning=F}
library(deSolve)
library(phaseR)
library(tidyverse)
```

Second let's write a convenience function that calls on phaseR under the hood to generate the flow field, nullclines, and sample trajectories for a given system and parameterization.

```{r}
phasePlot <- function(mod, params, xmax = 1, ymax = 1){
  xlim <- c(.1, xmax)
  ylim <- c(0, ymax)
  
  y0 <- matrix(c(1, 2),
             ncol = 2, nrow = 1,
             byrow = TRUE)  
  
  flw <- flowField(mod, xlim = xlim, ylim = ylim, parameters = params, 
                   xlab = 'Population', ylab = 'State coffers', add = F)
  nc <- nullclines(mod, xlim = xlim, ylim = ylim, parameters = params)
  trj <- trajectory(mod, y0 = y0, tlim = c(0,200), col = rep('black', nrow(y0)), parameters = params)
}
```


```{r}
time <- seq(from=0, to=500, by = 0.01)
 # parameters: a named vector
parameters <- c(r = .02, K = 100)
    
# initial conditions: also a named vector
state <- c(x = 10)

  logistic <- function(t, state, parameters){
        with(
            as.list(c(state, parameters)),{
                dx <- r*x*(1-x/K)
                return(list(dx))
            }
            )
  }
  
      out <- ode(y = state, times = time, func = logistic, parms = parameters)
ggplot(data = as.data.frame(out), aes(time, x)) + geom_point()
```
Lotka volterra
```{r}
# time sequence 
time <- seq(0, 50, by = 0.01)

# parameters: a named vector
parameters <- c(r = 2, k = 0.5, e = 0.1, d = 1)

# initial condition: a named vector
state <- c(V = 1, P = 3)

# R function to calculate the value of the derivatives at each time value
# Use the names of the variables as defined in the vectors above
lotkaVolterra <- function(t, state, parameters){
  with(as.list(c(state, parameters)), {
    dV = r * V - k * V * P
    dP = e * k * V * P - d * P
    return(list(c(dV, dP)))
  })
}

out <- ode(y = state, times = time, func = lotkaVolterra, parms = parameters)
    
out.df <- as.data.frame(out)

out.df %>%
  pivot_longer(-time) %>%
ggplot(aes(time, value, color = name)) + geom_point()


ggplot(out.df[1:567,], aes(P, V, color = time)) + geom_point()
```


# Model 1: Simple consumer-resource system with network effects
First, we'll replicate the model of Muneepeerakul and Qubbaj (2012). It's a simple consumer resource system, with parameterized flows of population and resources (i.e. immigration and trade).

Setup the model.

```{r}
aNt/(Nt+d)–Nt

tibble(n = 1:100) %>%
  mutate(test = 5 * n / (n+10) - n) %>%
  plot()
```


```{r}
nefedov <- function(t, y, parameters){
    r <- parameters[1]
    q <- parameters[2]
    N <- y[1]
    K <- y[2]

    dy <- numeric(2)
    dy[1] <- r * N * (1 - N / K)
    dy[2] <- q * 1 / (q - 1) * N / (N + 1 / (q - 1)) - N
    list(dy)
     
}
state_n <- c(N = .2, K = .2)

#a i sminimum food req? or numbers of min food req that can be gatehred
# oh a0 is average crop yield
# d is a random divisor
# total harvest is a_tN_t
out.nefedov <- ode(y = state_n, times =  seq(0, 1000, by = 1), func = nefedov, parms = c(0.02, 1.2))
    
out.df.nefedov <- as.data.frame(out.nefedov)

out.df.nefedov %>%
  pivot_longer(-time) %>%
ggplot(aes(time, value, color = name)) + geom_point()


ggplot(out.df.nefedov, aes(N, K, color = time)) + geom_point()
```

No scaling
```{r fig.width = 6, fig.height = 6}
phasePlot(nefedov, c(0.01, 1.2), xmax = 1.5, ymax = 1.5)

flowField(derivative, xlim = c(.1, 5), ylim = c(0, 5), parameters = c(0.02, 1.2), 
                   xlab = 'Population', ylab = 'State coffers', add = F)

phasePlaneAnalysis(netMod, c(.01,1.5), c(0,1.5), parameters = c(0.2, 1.2))
```


```{r}
plot.new()
trajectory(netMod, y0 = c(1,1), tlim = c(0, 10), parameters = c(0.01, 1.2))
```


Fiscal demographic model

```{r}
c <- 3
s0 <- 10
S <- 1:100
plot(1 + c * (S / (s0 + S)))

```


```{r}
# Define the Fiscal-Demographic model function
fiscal_demographic <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # Define the model equations
    K <- function(S) 1 + c * S / (s0 + S)
    dN <- r * N * (1 - N / K(S)) 
    dS <-  N * (1 - N / K(S)) - beta * N

    # Return the derivatives
    return(list(c(dN, dS)))
  })
}

# Define the initial conditions
N0 <- 0.5      # Initial population size [0-3]
S0 <- 0      # Initial state capacity [0-10]

# parameters: a named vector
parameters <- c(r = 0.015,    # growth rate [0-0.03]
                beta = 0.25, # per capita state expenditure [0-1]
                s0 = 10,# [0-1000]
                c = 3)   # Fiscal capacity coefficient [0-6]


# Create a vector of initial conditions
initial_state <- c(N = N0, S = S0)

posfun <- function(t, y, parms){
    with(as.list(y), {
      y[which(y<0)] <- 0  
      return(y)
    })
  }

# Define the time span
times <- seq(0, 500, by = 0.01)

# Solve the model equations
solution <- ode(y = initial_state, times = times, func = fiscal_demographic, parms = parameters,
                events=list(func = posfun, time = times))


solution %>%
  as.data.frame() %>%
  mutate(S = S / 30) %>%
  pivot_longer(-time)%>%
ggplot(aes(time, value, color = name)) +
  geom_line() +
  geom_hline(yintercept = c(1), linetype = 2)
```

```{r}
flowField(fiscal_demographic, xlim = c(0, 4), ylim = c(0, 50), parameters = parameters, 
                   xlab = 'Population', ylab = 'State coffers', add = F, state.names = c('N', 'S')) 
nullclines(fiscal_demographic, xlim = c(0, 4), ylim = c(0, 50), parameters = parameters, state.names = c('N', 'S'))
  y0 <- matrix(c(.5, 0, 1, 0, 0.1, 0),
             ncol = 2,
             byrow = TRUE)  
trajectory(fiscal_demographic, y0 = y0, tlim = c(0,500), col = rep('black', nrow(y0)), parameters = parameters, state.names = c('N', 'S'))
```

Internal warfare
```{r}
fiscal_demographic_war <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    K <- kmax - c * W

    dN <- r * N * (1 - N / K) - delta * N * W
    dS <- rho * N * (1 - N / K) - beta * N
    dW <- a * N ^ 2 - b * W - alpha * S

    # Return the derivatives
    return(list(c(dN, dS, dW)))
  })
}

parameters <- c(r = 0.015,    # growth rate [0-0.03]
                rho = 1,
                beta = 0.25, # per capita state expenditure [0-1]
                delta = 0.1,
                kmax = 3,
                a = 0.01,
                b = 0.05,
                c = 1, # not in paper
                alpha = 0.1)  

# Create a vector of initial conditions
initial_state <- c(N = .5, S = 0, W = .5)

posfun <- function(t, y, parms){
    with(as.list(y), {
      y[which(y<0)] <- 0  
      return(y)
    })
  }
# Define the time span
times <- seq(0, 1000, by = .01)

# Solve the model equations
solution <- ode(y = initial_state, times = times, func = fiscal_demographic_war, parms = parameters,
                events=list(func = posfun, time = times))


solution %>%
  as.data.frame() %>%
    mutate(S = S / 6) %>%
  pivot_longer(-time) %>%
  #filter(time > 500) %>%
ggplot(aes(time, value, color = name)) +
  geom_line()
```

warfare in stateless societies
```{r}
stateless_war <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # Define the model equations
    K <- 1# kmax - c * W
    dN <- r * N * (1 - N / K) - N * W
    dW <- (a * N ^ 2 - b * W)

    # Return the derivatives
    return(list(c(dN, dW)))
  })
}

parameters <- c(r = 0.015,    # growth rate [0-0.03]
                rho = 1,
                beta = 0.25, # per capita state expenditure [0-1]
                delta = 0.1,
                kmax = 3,
                a = 0.01,
                b = 0.05,
                c = 10, # not in paper
                alpha = 0.1)  

# Create a vector of initial conditions
initial_state <- c(N = .5, W = 0)


# Define the time span
times <- seq(0, 500, by = 0.01)

# Solve the model equations
solution <- ode(y = initial_state, times = times, func = stateless_war, parms = parameters,
                events=list(func = posfun, time = times))


solution %>%
  as.data.frame() %>%
  pivot_longer(-time)%>%
ggplot(aes(time, value, color = name)) +
  geom_line() +
  facet_wrap(~name, scales = 'free_y', ncol = 2)
```

Fiscal demographic model with class structure
```{r}
# Define the Fiscal-Demographic model function
fiscal_demographic_class <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # Define the model equations
    g <- g0 * (h + S) / (h + (1 + c) * S)
    dP <- (beta1 * rho0 * (1 - g * P) * P) / (1 + a * E) - delta1 * P
    dE <- (beta2 * rho0 * a * E * (1 - g * P) * P) / (1 + a * E) - delta2 * E

    dS <- gamma * dE - alpha * E

    # Return the derivatives
    return(list(c(dP, dE, dS)))
  })
}

# Define the initial conditions
P0 <- 0.5      # Initial population size [0-3]
S0 <- 0      # Initial state capacity [0-10]
E0 <- 0.2

# parameters: a named vector
parameters <- c(beta1 = 0.02,
                delta1 = 0.02,
                rho0 = 2,
                beta2 = 0.5,
                delta2 = 0.025,
                gamma = 10,
                g0 = 1,
                alpha = 0.2,
                c = 3, # ?
                h = 1, #?
                a = 1) #?   # Fiscal capacity coefficient [0-6]


# Create a vector of initial conditions
initial_state <- c(P = P0, S = S0, E = E0)

# Define the time span
times <- seq(0, 1000, by = 0.01)

# Solve the model equations
solution <- ode(y = initial_state, times = times, func = fiscal_demographic_class, parms = parameters,
                events=list(func = posfun, time = times))


solution %>%
  as.data.frame() %>%
  pivot_longer(-time)%>%
ggplot(aes(time, value, color = name)) +
  geom_line() +
  facet_wrap(~name, scales = 'free_y')
```

Selfish Elite model
```{r}
# Define the Fiscal-Demographic model function
selfish_elites <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # Define the model equations

    dP <- (beta1 * P * (1 - P)) / (1 + E) - delta1 * P
    dE <- (beta2 * E * (1 -  P) * P) / (1 + E) - delta2 / (1 + S) * E

    dS <- gamma * dE - alpha * E

    # Return the derivatives
    return(list(c(dP, dE, dS)))
  })
}

# Define the initial conditions
P0 <- .1      # Initial population size [0-3]
S0 <- 0      # Initial state capacity [0-10]
E0 <- .1

# parameters: a named vector
parameters <- c(beta1 = 0.02,
                delta1 = 0.02,
                rho0 = 2,
                beta2 = 0.5,
                delta2 = 0.025,
                gamma = 10,
                alpha = 0.2) 


# Create a vector of initial conditions
initial_state <- c(P = P0, E = E0, S = S0)

# Define the time span
times <- seq(0, 1000, by = 0.01)

# Solve the model equations
solution <- ode(y = initial_state, times = times, func = selfish_elites, parms = parameters,
                events=list(func = posfun, time = times))


solution %>%
  as.data.frame() %>%
  pivot_longer(-time)%>%
ggplot(aes(time, value, color = name)) +
  geom_line()
```
